Why do people punish? Evidence for a range of strategic concerns

### 2023-12-13 manuscript.Rmd
update manuscript

abstract: |
 Costly punishment is thought to be one of the key mechanisms sustaining cooperation in humans. However, the motives for punitive behaviour remain unclear. Punishment is often assumed to be motivated by a desire to convert free riders into cooperators, but it is also consistent with a host of other functions, such as levelling payoffs or increasing one’s relative position. We used a suite of six economic games to tease apart the different motives for punishment. Across representative samples from the United Kingdom and the United States, we estimated the frequency of different punishment strategies in the population, finding that egalitarian motives for punishment are more common than behaviour-change motives. Moreover, different punishment strategies were differentially predicted by personality, social preferences, political ideology, and religious views. Self-reports of behaviour in the games suggested that people have some degree of insight into their punishment strategy. These findings highlight the multipurpose nature of human punishment.

keywords          : "punishment; cooperation; economic games"
wordcount         : "5185 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : nature.csl
---

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One 
mechanism thought to sustain this level of cooperation is costly punishment, 
whereby individuals harm others at a personal cost[@CluttonBrock1995], 
ostensibly encouraging cooperative behaviour from the target (or bystanders [@DosSantos2011;@DosSantos2013;@Raihani2015]) in the future. Punishment 
therefore offers a route to maintaining or increasing cooperation by changing 
the payoff structure of social interactions such that it no longer pays to cheat 
or exploit social partners[@CluttonBrock1995;@Ostrom1990].

@@

Beyond behaviour-shaping concerns, there are a host of other reasons why people 
may want to punish in economic games. Punishers might be motivated by a desire 
for retribution rather than deterrence, punishing in proportion to the amount 
of harm that was personally caused[@Carlsmith2002]. Punishment might be driven 
by concerns about relative payoffs, such as disadvantageous inequity aversion 
(i.e., avoiding having less than others[@Raihani2012b;@Raihani2019]) and/or 
general egalitarian preferences (i.e., wanting all participants to receive the 
same payoffs[@Dawes2007]). Such concerns about relative payoffs may be activated 
when participants earn less than cheats in economic games or when there are 
income disparities in these settings. People might also use punishment for 
competitive purposes, seeking advantageous inequity for themselves (i.e., having 
more than others) and/or improving their relative position[@Raihani2019].

@@

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation) predict the use of different punishment strategies. Table 
\@ref(tab:tableStrategies) summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that Table \@ref(tab:tableStrategies) is not 
an exhaustive list of all possible punishment strategies: we do not include 
reputational functions of punishment in this table, such as signalling 
trustworthiness[@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], 
because our focus is on punishment strategies in anonymous economic games without 
reputational incentives (but see ref[@Jordan2020]).

@@

(ref:plotGamesCaption) *Visual summary of the six economic games.* In all
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does 
nothing. Player 1 is then given the option to punish by paying a certain amount 
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off by stealing, (B) Player 2 creates equality by stealing, (C) the
computer "decides" whether Player 2 steals, (D) the fee-fine ratio is 1:1, (E)
Player 2 is better off by stealing, and (F) Player 2 steals instead from a 
third-party.

@@

The overall pattern of punitive behaviour in the six economic games was in line 
with previous research and very similar across both countries (Figure 
\@ref(fig:plotDecisions)). Participants were generally more likely to punish 
targets who stole from another individual compared to targets who did not steal 
(multilevel logistic regression; _b_ = 
`r printnum(mlm1["otherTook","Estimate"])`, standard error = 
`r printnum(mlm1["otherTook","Std. Error"])`, _p_ 
`r printp(mlm1["otherTook","Pr(>|z|)"])`). Participants were also more likely 
to punish when targets' stealing behaviour generated inequalities, 
specifically in Games E and F (_b_ = 
`r printnum(mlm2["DI","Estimate"])`, SE = `r printnum(mlm2["DI","Std. Error"])`, 
_p_ `r printp(mlm2["DI","Pr(>|z|)"])`).

@@

We classified participants into a particular strategy if their behaviour across 
all twelve decisions matched our behavioural predictions shown in Table 
\@ref(tab:tableStrategies) exactly. Table \@ref(tab:tableStrategyCounts) shows 
the proportion of participants following each strategy, with N/A used to 
represent participants who did not fit exactly into any particular strategy type.
Overall, `r printnum(mean(tar_read(dExc)$strategy != "N/A")*100, digits = 0)`%
of our participants could be classified exactly into one of the strategies. The
most common strategy in both countries was to never punish across any of the
games. The next most common strategies were those that care about
minimising payoff differences (avoid disadvantageous inequity, egalitarian). 
Less common were the behaviour-shaping strategies (deterrent, norm-enforcing),
the retributive strategy, and the competitive strategies (seek AI, competitive). 
Although participants often punished targets who did not steal 
in the six games (Figure \@ref(fig:plotDecisions)), no participants followed the 
antisocial strategy by exclusively punishing targets who did not steal across 
_all_ games.

(ref:tableStrategyCountsCaption) Counts and proportions of participants 
following each punishment strategy exactly, split by country. N/A implies that
participants were unable to be classified exactly into any of the punishment
strategies.

@@

The posterior estimates from the model are presented in Figure 
\@ref(fig:plotModel1b). The posterior probabilities for each strategy did not
differ between the two countries. In both countries, the never punish strategy
had the highest probability, followed by the egalitarian strategy. The
norm-enforcing and seek AI strategies were the next most likely, with higher
posterior estimates than the competitive and antisocial strategies. None of the
other strategies differed in their posterior estimates. The same general pattern
emerged when we analysed the full dataset without pre-registered exclusions 
(Supplementary Figure \@ref(fig:plotModel1a)).

(ref:plotModel1bCaption) *Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model.* The model 
assumes an implementation error rate of 5%. Points represent posterior medians, 
line ranges represent 50% and 95% credible intervals.

@@

Conversely, personality and social preferences were linked to variation in 
punishment strategies. When including the Big-6 personality dimensions and 
Social Value Orientation (SVO) in the model, we found associations with the 
egalitarian, never punish, and random choice strategies (Figure
\@ref(fig:plotAllPers2)). Participants higher in SVO were more likely to follow 
the egalitarian and the never punish strategies, while those with lower SVO 
scores were more likely to enact the random choice strategy. The personality 
dimensions of honesty-humility and openness to experience were both positively 
associated with following the never punish strategy, while extraversion 
negatively predicted this strategy. The effects were mostly similar across
countries, but occasionally differed: for example, in the United States,
but not in the United Kingdom, honesty-humility was positively associated with 
following the egalitarian strategy and negatively associated with following 
the random choice strategy. Overall, the same pattern of results emerged when 
analysing the full dataset without exclusions (Supplementary Figure 
\@ref(fig:plotAllPers1)).

(ref:plotAllPers2Caption) *Posterior slopes from Bayesian latent state models 
including Big-6 personality dimensions and Social Value Orientation.* Each row 
represents a separate model. All models assume an implementation error rate of 
5%. Points represent posterior medians, line ranges represent 95% credible 
intervals.

@@

Political and religious variables were also associated with punishment strategy 
(Figure \@ref(fig:plotAllPolRel2)). These effects tended to be more pronounced 
in the United States. Controlling for Social Dominance Orientation, American 
participants higher in Right Wing Authoritarianism were more likely to follow 
the strategies avoiding disadvantageous inequity and seeking advantageous 
inequity. Participants who stated that they would like to "bring those below 
them [on the socio-economic status ladder] up a peg" were more likely to follow 
the egalitarian strategy, while American participants higher in Social Dominance 
Orientation, Right Wing Authoritarianism, and believing that God controls events 
in the world were less likely to follow the egalitarian strategy. In general, 
religious and conservative participants were less likely to follow the never 
punish strategy. This general pattern of results was replicated with the full 
dataset (Supplementary Figure \@ref(fig:plotAllPolRel1)).

(ref:plotAllPolRel2Caption) *Posterior slopes from Bayesian latent state models 
including political ideology, views about social inequality, and religiosity.* 
Each row represents a separate model aside from Social Dominance Orientation and 
Right Wing Authoritarianism, which control for one another within the same 
model. All models assume an implementation error rate of 5%. Points represent 
posterior medians, line ranges represent 95% credible intervals.

@@

In general, we found that self-reported strategy usage was positively associated
with the behavioural strategy that participants employed (see Supplementary 
Figures \@ref(fig:plotSliders1) and \@ref(fig:plotSliders2) for the distribution 
of responses to self-report questions). Figure \@ref(fig:plotAllSliders2) shows 
the relationships between self-report questions and the different punishment 
strategies, highlighting the combinations where the question matched the 
behavioural strategy. We found positive relationships between the self-report 
questions and strategy usage for the norm-enforcing, egalitarian, seek 
advantageous inequity, never punish, and random choice  strategies. The 95% 
credible intervals for other estimates included zero, though these estimates 
often trended in a positive direction. The same pattern of results was found 
when analysing the full dataset without exclusions  (Supplementary Figure 
\@ref(fig:plotAllSliders1)).

(ref:plotAllSliders2Caption) *Posterior slopes from models including 
self-reported strategy usage.* Each row represents a separate model. All models 
assume an implementation error rate of 5%. Highlighted estimates represent 
combinations where the self-report question matched the behavioural strategy. 
Points represent posterior medians, line ranges represent 95% credible 
intervals.

@@

Using a suite of economic games measuring punishment in different situations, 
we have shown that punishment does not serve just one function, but instead is 
a flexible tool that can be and is used for different purposes[@Raihani2019]. 
Punishment is more akin to a swiss army knife than a hammer, used by some to 
enforce norms of cooperation and by others to reduce or even create inequality 
between individuals. We found that people's punishment strategy can, to some 
extent, be predicted by individual differences in personality, social 
preferences, and political and religious views. Moreover, contrary to the view 
that people are often unable to articulate the reasons for their punitive 
behaviour[@Carlsmith2002], people seem to have some degree of insight into the 
strategy they are using. Despite small differences, these general patterns 
replicated in samples from both the United Kingdom and the United States.

Among the punitive strategies, the most common were particularly sensitive to 
inequality in payoffs, either from a self-referential perspective (i.e., avoid 
disadvantageous inequity) or more generally (i.e., egalitarian). This is in 
line with previous studies which have highlighted inequity aversion as an 
important motivation for punishment in economic games 
[@Dawes2007;@Raihani2012b]. Personality and social preference variables mapped 
onto these strategies in expected ways. Traits associated with other-regarding 
concern, such as SVO and honesty-humility, predicted following the egalitarian 
strategy, whereas religious and conservative individuals were less likely to 
follow this strategy, especially in the United States. Moreover, participants 
following the egalitarian strategy were able to self-report this strategy, 
though the same was not true for the avoid disadvantageous inequity strategy.

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less 
common than strategies sensitive to inequality in our set of games. This was 
reflected both in participants’ elicited punishment behaviour (Figure 
\@ref(fig:plotModel1b)) and in their self-reports of their own strategy 
(Supplementary Figures \@ref(fig:plotSliders1) and \@ref(fig:plotSliders2)). 
Regarding the predictors of these strategies, we found that demographic, 
personality, political, and religious variables tended to be unrelated to 
deterrent and norm-enforcing punishment strategies. We also found that 
participants had insight into the norm-enforcing strategy, but not the deterrent 
strategy. This finding is in line with previous research showing that people 
struggle to accurately report the deterrent motivations for their punitive 
behaviour[@Carlsmith2002].

Other punitive strategies were less common in our dataset, but some were more 
prevalent than others. For example, participants were more likely to use 
punishment to seek advantageous inequity than to exclusively harm those who did 
not steal (i.e., antisocial punishment). The existence of the "seek AI" strategy 
in our dataset supports the claim that punishment can also be used as a tool to 
increase one’s own relative position[@Raihani2019]. While generally rare, we 
found that this motive for punishment was more common among authoritarian 
participants, at least in the United States, potentially providing an 
explanation for why peer punishment has been found to be more common among 
conservatives in previous work[@Hofmann2018]. Moreover, the fact that no 
participants in our sample punished non-stealing across all games suggests that 
antisocial punishment does not function to harm cooperators specifically, as has 
been previously suggested[@Herrmann2008]. Instead, antisocial punishment appears 
to be motivated by improving one's relative position in general, which is in 
line with work showing that antisocial punishment disappears with a 1:1 fee-fine 
ratio[@Sylwester2013].

The fact that people use punishment for many different reasons poses problems 
for the way that punishment is operationalised in classic behavioural economic 
game studies. In these studies, a common assumption is that participants will 
punish to change the behaviour of cheats. But in reality, people may be choosing 
the punishment option to achieve a variety of different goals. The targets of 
punishment in these studies are likely well aware that punishment could be 
levied for these different reasons and this knowledge may impact their 
responses. For example, if cheating targets interpret punishment as serving a 
competitive motive, it may elicit retaliation rather than encourage 
cooperation[@Dreber2008;@Nikiforakis2008b;@Raihani2019]. This might help to 
explain the mixed findings in the field as to whether punishment actually 
motivates cheating targets to cooperate in the future[@Raihani2019].

It is striking that the most common strategy in our dataset was to never punish. 
This is partly because punishment in these games imposes an economic cost for no 
tangible benefit. If the fee-fine ratio had been lower such that it was cheaper 
to punish, we may have seen more punishment from participants. Indeed, 
`r printnum(mean(tar_read(dExc)$SelfRate10[tar_read(dExc)$strategy == "Never punish"] > 0, na.rm = TRUE)*100, digits = 0)`% 
of participants following the never punish strategy positively stated that they 
didn't want to pay to reduce anyone's bonus but would have done so if it were 
free. But the frequency of the never punish strategy perhaps also reflects a 
more general aversion to peer punishment, an aversion that has been highlighted 
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples
[@Balafoutas2014;@Balafoutas2016] and in small-scale societies[@Baumard2010]. 
One reason that people may be averse to peer punishment is that, due to its 
multipurpose nature, it may be interpreted as a competitive challenge by targets 
and trigger retaliation[@Raihani2019]. In situations that lack clear 
institutional norms to legitimise punishment, such as our economic games and 
some situations in the real world, people might abstain from peer punishment to 
avoid such retaliation, regardless of whether retaliation is actually possible. 
By contrast, institutionalised punishment in small-scale societies often 
functions to compensate victims adequately while limiting the potential for 
feuds and cycles of retaliation[@Fitouchi2023;@Singh2022]. Future research 
should uncover whether people are more willing to punish in these 
conventionalised contexts.

There are several limitations with our study design that can guide future 
research. First, we used one-shot economic games to measure punishment 
strategies, which may have led us to underestimate behaviour-change strategies 
like deterrence. Our inclusion of Game C somewhat mitigated this issue by 
manipulating whether stealing behaviour was intentional vs. unintentional and 
thus whether there was any behaviour to be deterred. Due to limits on our 
within-subjects design and the complexity of the strategy space, it was not 
feasible for us to expand our study to include additional contexts to elicit 
behaviour-change strategies (e.g., repeated games, games where targets are not 
made aware of the punishment, games where targets can retaliate). Future work 
could study these contexts separately.

>>>>

abstract: |
 Costly punishment is thought to be a key mechanism sustaining human cooperation. However, the motives for punitive behaviour remain unclear. Although often assumed to be motivated by a desire to convert cheats into cooperators, punishment is also consistent with other functions, such as levelling payoffs or improving one’s relative position. We used six economic games to tease apart different motives for punishment and to explore whether different punishment strategies were associated with personality variables, political ideology, and religiosity. We used representative samples from the United Kingdom and the United States (N = 2010) to estimate the frequency of different punishment strategies in the population. The most common strategy was to never punish: prosocial individuals and those scoring high in honesty-humility were among the least punitive in our sample, while religious individuals and those scoring high for right-wing authoritarianism and social dominance orientation were more punitive. For people who did punish, strategy use was more consistent with egalitarian motives than behaviour-change motives. Nevertheless, different punishment strategies were also associated with personality, social preferences, political ideology, and religiosity. Self-reports of behaviour in the games suggested that people have some insight into their punishment strategy. These findings highlight the multipurpose nature of human punishment and show how the different motives underpinning punishment decisions are linked with core character traits.

keywords          : "punishment; cooperation; economic games"
wordcount         : "xxxx words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : nature.csl
---

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One 
mechanism thought to sustain this level of cooperation is costly punishment, 
whereby individuals harm others at a personal cost[@CluttonBrock1995], 
ostensibly encouraging cooperative behaviour from the target (or bystanders 
[@DosSantos2011;@DosSantos2013;@Raihani2015]) in the future. Punishment offers a 
route to maintaining or increasing cooperation by changing the payoff structure 
of social interactions such that it no longer pays to cheat or exploit social 
partners[@CluttonBrock1995;@Ostrom1990]. Yet, despite its theoretical 
importance, the question of why people choose to punish others is still hotly 
contested[@Raihani2019]. In this study, we use a battery of economic games to 
disentangle the different motives underpinning punishment and explore how these 
motives vary across individuals.

@@

Beyond behaviour shaping concerns, there are a host of other reasons why people 
may want to punish in economic games. Punishers might be motivated by a desire 
for retribution rather than deterrence, punishing in proportion to the amount 
of harm that was personally caused[@Carlsmith2002]. Punishment might be driven 
by concerns about relative payoffs, such as disadvantageous inequity aversion 
(i.e., avoiding having less than others[@Raihani2012b;@Raihani2019]) and/or 
general egalitarian preferences (i.e., wanting all participants to receive the 
same payoffs[@Dawes2007]). Such concerns about relative payoffs may be activated 
when participants earn less than cheats in economic games or when there are 
income disparities in these settings. People might also use punishment for 
competitive purposes, seeking advantageous inequity for themselves (i.e., having 
more than others) and/or improving their relative position[@Raihani2019].

@@

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation, religious views) predict the use of different punishment strategies.
Table \@ref(tab:tableStrategies) summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that Table \@ref(tab:tableStrategies) is not 
an exhaustive list of all possible punishment strategies: we do not include 
reputational functions of punishment in this table, such as signalling 
trustworthiness[@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], 
because our focus is on punishment strategies in anonymous economic games without 
reputational incentives (but see ref[@Jordan2020]).

@@

(ref:plotGamesCaption) *Visual summary of the six economic games.* In all
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does 
nothing. Player 1 is then given the option to punish by paying a certain amount 
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off after stealing, (B) Player 2 creates equality by stealing, (C) 
the computer "decides" whether Player 2 steals, (D) the fee-fine ratio is 1:1, 
(E) Player 2 is better off after stealing, and (F) Player 2 steals instead from 
a third-party.

@@

The overall pattern of punitive behaviour in the six economic games was very 
similar across both countries (Figure \@ref(fig:plotDecisions)). Participants 
were generally more likely to punish targets who stole compared to targets who 
did not steal (multilevel logistic regression; _b_ = 
`r printnum(mlm1["otherTook","Estimate"])`, standard error = 
`r printnum(mlm1["otherTook","Std. Error"])`, _p_ 
`r printp(mlm1["otherTook","Pr(>|z|)"])`). Participants were also more likely 
to punish when targets' stealing behaviour generated inequalities, 
specifically in Games E and F (_b_ = 
`r printnum(mlm2["DI","Estimate"])`, SE = `r printnum(mlm2["DI","Std. Error"])`, 
_p_ `r printp(mlm2["DI","Pr(>|z|)"])`).

@@

We classified participants into a particular strategy if their behaviour across 
all twelve decisions matched our behavioural predictions shown in Table 
\@ref(tab:tableStrategies) exactly. Table \@ref(tab:tableStrategyCounts) shows 
the proportion of participants following each strategy, with N/A used to 
represent participants who did not fit exactly into any particular strategy type.
Overall, `r printnum(mean(tar_read(dExc)$strategy != "N/A")*100, digits = 0)`%
of our participants could be classified exactly into one of the strategies. The
most common strategy in both countries was to never punish across any of the
games. The next most common strategies were those that care about
minimising payoff differences (avoid disadvantageous inequity, egalitarian). 
Less common were the behaviour shaping strategies (deterrent, norm-enforcing),
the retributive strategy, and the competitive strategies (seek advantageous 
inequity, competitive). Although participants often punished targets who did not 
steal in the six games (Figure \@ref(fig:plotDecisions)), no participants 
followed the antisocial strategy by exclusively punishing targets who did not 
steal across _all_ games.

(ref:tableStrategyCountsCaption) Counts and proportions of participants 
following each punishment strategy exactly, split by country. *N/A implies that
participants were unable to be classified exactly into any of the punishment
strategies. AI = advantageous inequity, DI = disadvantageous inequity.*

@@

The posterior estimates from the model are presented in Figure 
\@ref(fig:plotModel1b). The posterior probabilities for each strategy did not
differ between the two countries. In both countries, the never punish strategy
had the highest probability, followed by the egalitarian strategy. The
norm-enforcing and seek advantageous inequity strategies were the next most 
likely, with higher posterior estimates than the competitive and antisocial 
strategies. None of the other strategies differed in their posterior estimates. 
The same general pattern emerged when we analysed the full dataset without 
pre-registered exclusions (Supplementary Figure \@ref(fig:plotModel1a)).

(ref:plotModel1bCaption) *Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model.* The model 
assumes an implementation error rate of 5%. Points represent posterior medians, 
line ranges represent 50% and 95% credible intervals. AI = advantageous 
inequity, DI = disadvantageous inequity.

@@

Conversely, personality and social preferences were linked to variation in 
punishment strategies. When including the Big-6 personality dimensions and 
Social Value Orientation (SVO) in the model, we found associations with the 
egalitarian, never punish, and random choice strategies, with small-to-medium 
effect sizes (Figure \@ref(fig:plotAllPers2)). More prosocial participants 
(those with higher SVO scores) were more likely to follow the egalitarian and 
the never punish strategies, while those with lower SVO scores were more likely 
to enact the random choice strategy. The personality dimensions of 
honesty-humility and openness to experience were both positively associated with 
following the never punish strategy, while extraversion negatively predicted 
this strategy. The effects were mostly similar across countries, but 
occasionally differed: for example, in the United States, but not in the United 
Kingdom, honesty-humility was positively associated with following the 
egalitarian strategy and negatively associated with following the random 
choice strategy. Overall, the same pattern of results emerged when analysing the 
full dataset without exclusions (Supplementary Figure \@ref(fig:plotAllPers1)).

(ref:plotAllPers2Caption) *Posterior slopes from Bayesian latent state models 
including Big-6 personality dimensions and Social Value Orientation.* Each row 
represents a separate model. All models assume an implementation error rate of 
5%. Points represent posterior medians, line ranges represent 95% credible 
intervals. AI = advantageous inequity, DI = disadvantageous inequity.

@@

Political and religious variables were also associated with punishment strategy 
(Figure \@ref(fig:plotAllPolRel2)). These effects were small-to-medium in size 
and tended to be more pronounced in the United States. Controlling for Social 
Dominance Orientation, American participants higher in Right Wing 
Authoritarianism were more likely to follow the strategies avoiding 
disadvantageous inequity and seeking advantageous inequity. Participants who 
stated that they would like to "bring those below them [on the socio-economic 
status ladder] up a peg" were more likely to follow the egalitarian strategy, 
while American participants higher in Social Dominance Orientation, Right Wing 
Authoritarianism, and believing that God controls events in the world were less 
likely to follow the egalitarian strategy. In general, religious and 
conservative participants were less likely to follow the never punish strategy. 
This general pattern of results was replicated with the full dataset 
(Supplementary Figure \@ref(fig:plotAllPolRel1)).

(ref:plotAllPolRel2Caption) *Posterior slopes from Bayesian latent state models 
including political ideology, views about social inequality, and religiosity.* 
Each row represents a separate model aside from Social Dominance Orientation and 
Right Wing Authoritarianism, which control for one another within the same 
model. All models assume an implementation error rate of 5%. Points represent 
posterior medians, line ranges represent 95% credible intervals. AI = 
advantageous inequity, DI = disadvantageous inequity.

@@

In general, we found that self-reported strategy usage was positively associated
with the behavioural strategy that participants employed, with effects ranging 
from small to large in size (see Supplementary Figures \@ref(fig:plotSliders1) 
and \@ref(fig:plotSliders2) for the distribution of responses to self-report 
questions). Figure \@ref(fig:plotAllSliders2) shows the relationships between 
self-report questions and the different punishment strategies, highlighting the 
combinations where the question matched the behavioural strategy. We found 
positive relationships between the self-report questions and strategy usage for 
the norm-enforcing, egalitarian, seek advantageous inequity, never punish, and 
random choice  strategies. The 95% credible intervals for other estimates 
included zero, though these estimates often trended in a positive direction. The 
same pattern of results was found when analysing the full dataset without 
exclusions  (Supplementary Figure \@ref(fig:plotAllSliders1)).

(ref:plotAllSliders2Caption) *Posterior slopes from models including 
self-reported strategy usage.* Each row represents a separate model. All models 
assume an implementation error rate of 5%. Highlighted estimates represent 
combinations where the self-report question matched the behavioural strategy. 
Points represent posterior medians, line ranges represent 95% credible 
intervals. AI = advantageous inequity, DI = disadvantageous inequity.

@@

Using a suite of economic games measuring punishment in different situations, 
we have shown that punishment does not serve just one function, but instead is 
a flexible tool that can be and is used for different purposes[@Raihani2019]. 
While some use punishment to enforce norms of cooperation, others use it to 
reduce or even create inequality between individuals. We found that people's 
punishment strategy can, to some extent, be predicted by individual differences 
in personality, social preferences, and political and religious views. Moreover, 
contrary to the view that people are often unable to articulate the reasons for 
their punitive behaviour[@Carlsmith2002], people seem to have some degree of 
insight into the strategy they are using. Despite small differences, these 
general patterns replicated in samples from both the United Kingdom and the 
United States, providing further confidence in the results.

Among the punitive strategies, the most common were particularly sensitive to 
inequality in payoffs, either from a self-referential perspective (i.e., avoid 
disadvantageous inequity) or more generally (i.e., egalitarian). This is in 
line with previous studies which have highlighted inequity aversion as an 
important motivation for punishment in economic games 
[@Bone2015;@Bone2016;@Dawes2007;@Deutchman2021;@Johnson2009;@Raihani2012b]. 
Personality, social preferences, and political and religious views were related 
to these strategies. Traits associated with other-regarding concern, such as SVO 
and honesty-humility, predicted following the egalitarian strategy. By contrast,
religious and conservative individuals, including individuals higher in SDO and
RWA, were less likely to follow the egalitarian strategy and more likely to 
follow the avoid disadvantageous inequity strategy, especially in the United 
States. Authoritarian participants in the United States were also more likely to 
actively seek advantageous inequity with their punishment. These findings 
provide insight into the motives that may have driven higher levels of peer 
punishment among conservatives in previous work[@Hofmann2018].

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less 
common than strategies sensitive to inequality in our set of games. This was 
reflected both in participants’ elicited punishment behaviour (Figure 
\@ref(fig:plotModel1b)) and in their self-reports of their own strategy 
(Supplementary Figures \@ref(fig:plotSliders1) and \@ref(fig:plotSliders2)). 
Although our design did not explicitly allow for behaviour shaping as the 
interactions were all one-shot, we did manipulate whether the target's stealing 
behaviour was intentional or not (Game C), an approach which has been used in 
previous vignette studies to identify behaviour shaping motives[@Carlsmith2002].
The lower prevalence of behaviour shaping strategies in our study is consistent 
with prior work showing that punishment often continues to be used when 
behaviour shaping is impossible, such as when the target will never find out 
that they have been punished[@Crockett2014] or on the last round of repeated 
interactions[@Barclay2016]. We found that demographic, personality, political, 
and religious variables tended to be unrelated to behaviour shaping strategies, 
though this could reflect low power to detect such associations given the low 
prevalence of these strategies in our sample. We also found that participants 
accurately reported using the norm-enforcing strategy, but not the deterrent 
strategy. This finding is in line with previous research showing that people 
struggle to accurately report the deterrent motivations for their punitive 
behaviour[@Carlsmith2002].

We defined antisocial punishment as occurring when a participant exclusively 
punished those who did not steal. There were no participants in our sample who 
followed this strategy, although some players did punish non-stealing
co-players. Harming non-stealing individuals was also consistent with the 
competitive strategies, which did appear in our sample albeit at low frequencies.
The fact that no participants in our sample punished non-stealing across all 
games suggests that the antisocial punishment that has been observed in other 
studies[@Herrmann2008;@Pleasant2018] is not aimed at harming cooperators 
specifically, as has been previously suggested[@Herrmann2008]. Instead, 
antisocial punishment is more likely to be motivated by improving one's 
relative position, which is in line with work showing that antisocial punishment 
disappears when relative payoffs cannot be changed (e.g., when punishment is 
only available with a 1:1 fee-fine ratio[@Sylwester2013]).

The fact that people use punishment for many different reasons poses problems 
for the way that punishment is operationalised in classic behavioural economic 
game studies. In these studies, a common assumption is that participants punish 
to change the behaviour of cheats[@Fehr2000;@Fehr2002]. But in reality, people 
may be choosing the punishment option to achieve a variety of different goals. 
This has implications for how people respond to being targeted by punishers in 
these games. Targets of punishment in these studies may know that punishment 
reflects different motives and can respond accordingly. For example, if targets
interpret punishment as serving a competitive motive, it may elicit retaliation
rather than encourage cooperation[@Dreber2008;@Nikiforakis2008b;@Raihani2019].
As punishers’ motives must be inferred (and such inferences likely depend on 
character traits of the target, as well as the context in which punishment 
occurs), there is likely to be some variation and error in attributing motives 
to punishers. To the extent that inferred motives affect target responses, this 
might help to explain the mixed findings in the field as to whether punishment 
actually motivates cheating targets to cooperate in the future[@Raihani2019].

It is striking that the most common strategy in our dataset was to never punish. 
This is partly because punishment in these games imposes an economic cost for no 
tangible benefit. If the fee-fine ratio had been lower, such that it was cheaper 
to punish, we may have seen more punishment from participants. Indeed, 
`r printnum(mean(tar_read(dExc)$SelfRate10[tar_read(dExc)$strategy == "Never punish"] > 0, na.rm = TRUE)*100, digits = 0)`% 
of participants following the never punish strategy positively stated that they 
didn't want to pay to reduce anyone's bonus but would have done so if it were 
free. But the frequency of the never punish strategy perhaps also reflects a 
more general aversion to peer punishment, an aversion that has been highlighted 
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples
[@Balafoutas2014;@Balafoutas2016] and in small-scale societies[@Baumard2010]. 
One reason that people may be averse to peer punishment is that, because it is 
a fundamentally harmful act, punishment can reflect badly on the punisher and 
people might therefore refrain from punishing others to avoid reputational 
damage[@Raihani2015]. People frequently avoid taking actions that could harm 
their reputation, even when they don’t know if reputation is at stake (as in the 
one-shot anonymous settings used here). Another reason that people may be averse
to peer punishment is that it can trigger retaliation[@Raihani2019]. This may be
especially likely in situations that lack clear institutional norms to 
legitimise punishment, such as our economic games. As with reputation damage, 
people might abstain from peer punishment to avoid retaliation, regardless of 
whether retaliation is actually possible. By contrast, institutionalised 
punishment in small-scale societies often functions to compensate victims while 
limiting the potential for feuds and cycles of retaliation
[@Fitouchi2023;@Singh2022]. Future research should uncover whether people are 
more willing to punish in these conventionalised contexts (e.g., see
[@Molleman2019]).
